---
title: "Extrem events vs. forest GPP"
author: "liang chen"
date: "26.2.2021"
output:
  rmdformats::readthedown:
    use_bookdown: true
    lightbox: true
    highlight: tango
---

```{r,warning=FALSE}
library(tidyverse)
library(RSQLite)
library(ggmap)
library(lubridate)
library(SPEI)
library(optimx)
library(lme4)
library(parameters)
library(sjPlot)
library(sjmisc)
library(reticulate)
library(stargazer)
library(leaflet)
library(dlnm)
library(fitdistrplus)
# give the path of python to R
use_python("C:\\Users\\liangch\\Anaconda3\\python.exe")
```
```{r setup, include=FALSE,echo=FALSE}
knitr::opts_chunk$set(python.reticulate = F)
```

### Link to database

```{r}

forest <- read.table("../ForestSite.txt",header = T,sep = ",")

conn <- dbConnect(SQLite(),"../Flux40SiteDataset.db")
```

In order to, reduce the cpu pressure, we prefure slicing at database by sql first then loading to R...

### Tmax and Tmin for SPEI drought index

```{r,message=FALSE,warning=FALSE}

site_list <- forest$SITE_ID
Tem <- NULL
for (s in site_list) {
  res <-
    dbGetQuery(
      conn,
      'SELECT TIMESTAMP_END, TA_F, site FROM Flux44SiteHalfHourly WHERE site = ?',
      params = s
    )
  res <- data.frame(res)
  res$TIMESTAMP_END <- ymd_hm(res$TIMESTAMP_END)  # to date
  res$year <- year(res$TIMESTAMP_END)
  res$month <- month(res$TIMESTAMP_END)
  res$day <- day(res$TIMESTAMP_END)
  
  nd <- res %>% group_by(site, year, month, day) %>%
    summarise(Tmax = max(TA_F), T_min = min(TA_F))
  Tem <- rbind(Tem, nd)
  messages <- cat("now is working on", s)
  print(messages)
}

```
### Prepare for SPEI drught index

**SPEI** is calculated from climatic water balance (precipitation minus potential evapotranspiration) at *month* scale. Thus we should get the water balance before hand. It also should be noticed that operation for precipitation is summation.

In addition,  PET for each month at each site is calculate from **thornthwaite** function. one of parameters is latitude information.

```{r LinkToDB, echo=T,warning=F}

ClimaticData <-
  dbGetQuery(conn,
             'SELECT TIMESTAMP,site,TA_F, P_F, Year,Month, Day FROM Flux44SiteDaily')
ClimaticData <- data.frame(ClimaticData)
# re-order the levels
ClimaticData$Month <-
  factor(ClimaticData$Month, levels =  c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12))
# Precipitation
Preci <- ClimaticData[, c("Year", "Month", "Day", "P_F", "site")]

# monthly precipitation
monthly_preci <- Preci %>%
  group_by(site, Year, Month) %>%
  summarise(Pre = sum(P_F))
# Other climatic variables
monthly_Tem <- ClimaticData[, c("Year", "Month", "TA_F", "site")] %>%
  group_by(site, Year, Month) %>%
  summarise(Ta_m = mean(TA_F))

# merge latitude and monthly temerature and precipitation
MonthlyData <- merge(monthly_preci, monthly_Tem,
                     by = c("site", "Year", "Month")) %>%
  merge(forest[, c("SITE_ID", "LOCATION_LAT")],
        by.y = "SITE_ID",
        by.x =  "site")
str(MonthlyData)
```

#### Calculations of PET 
```{r PET, message=F,warning=F}

site_list  <- unique(MonthlyData$site)
PET <- NULL
sites <- NULL
for (s in site_list) {
  siteData <- subset(MonthlyData, site == s)
  years <- unique(siteData$Year)
  
  for (y in years) {
    subd <- subset(siteData, Year == y)
    sites <- append(sites, s)
    PE <- thornthwaite(ts(
      subd$Ta_m,
      start = c(subd$Year[1], 1),
      frequency = 12
    ),
    subd$LOCATION_LAT[1])
    PET <- rbind(PET, PE)
  }
}
```

#### Calculations of SPEI
```{r,fig.align='center'}
MonthlyData$PET  <- PET
MonthlyData$Bal <- MonthlyData$Pre-MonthlyData$PET
MonthlyData <- na.omit(MonthlyData)
# 1 year resolution
spei1 <- spei(MonthlyData$Bal,1)
plot(spei1)
```

#### SPEI site-wise distribution
```{r,fig.height=20,fig.width=20,fig.cap="Site-wise SPEI distribution",out.height=25,out.width=25, warning=FALSE}
MonthlyData$spei <- spei1$fitted
MonthlyData %>% group_by(site,Year) %>%
summarise(spei_m = mean(spei)) %>%
ggplot(aes(factor(Year),spei_m))+geom_point(col="red")+
    facet_wrap(site~.,ncol = 5,scale="free")+
    theme(axis.text.x = element_text(angle = 90))
```
#### transform to yearly
```{r, warning=FALSE}
yearlyPreci <- MonthlyData %>%
  group_by(site, Year) %>%
  summarise(Pre_total = sum(Pre))

drought_yearly <- MonthlyData %>% group_by(site, Year) %>%
  summarise_all(mean)

drought_yearly$sumPreci <- yearlyPreci$Pre_total
```

### Frost
the frost period is varied from year to year, due to we adjusted the method.
The current method is based on Carbon flux phenology (GPP phenology). Basic description for this method as follow:

- the first time in a year when GPP = 10% GPPmax, the day we defined as the start of growing season.

- the second time in a year when GPP = 10% GPPmax, the day we defined as the end of growing season.

This part was done using python, code as follow:

```python
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import sqlite3
from statsmodels.tsa.holtwinters import ExponentialSmoothing, SimpleExpSmoothing, Holt

conn1 = sqlite3.connect('../Flux40SiteDataset.db')

md = pd.read_sql('select * from Flux44SiteDaily', conn1)

md = pd.DataFrame(md)
md = md.loc[md['GPP_NT_VUT_REF']!= -9999.000000,] #-9999 means NA

site_list = np.unique(md['site'].values)

smoothedGPP = []

for site in site_list:
    sub = md[md["site"] == site]
    years = np.unique(sub['Year'])
    
    for year in years:
    #   
        subb = sub[sub["Year"] == year]
        
    # 
        fit1 = SimpleExpSmoothing(subb['GPP_NT_VUT_REF'].values).fit(smoothing_level=0.15, optimized=False)
     
        subb["smoothedGPP"] = fit1.fittedvalues
    # 
        smoothedGPP.append(subb)
    print(f"now is {site}")
  
smoothedGPP = pd.concat(smoothedGPP)


# calculation for SOG AND EOG
phe_gpp = pd.DataFrame(columns=["site", "Year", "Month", "Day", "smoothedGPP", "GPP_f", "datetime"])
# phe_nep = pd.DataFrame(columns=["year", "DOY"])
for site in site_list:
    sub = smoothedGPP[smoothedGPP["site"] == site]
    years = np.unique(sub["Year"])
    for year in years:
        subb = sub[sub['Year'] == year]
        #     phe_yearly = Phe_Hyy[Phe_Hyy['Year'] == year]
        print(f'Now calculating the {year} at {site}')
        for i in np.arange(len(subb)):

            if subb['smoothedGPP'].values[i] >= 0.1 * max(subb['GPP_NT_VUT_REF'].values):
                phe_gpp = phe_gpp.append({"site": site, "Year": year, "Month": subb["Month"].values[i],
                                          "Day": subb["Day"].values[i],
                                          "smoothedGPP": subb["smoothedGPP"].values[i],
                                          "GPP_f": subb["GPP_NT_VUT_REF"].values[i],
                                          "datetime": subb["TIMESTAMP"].values[i]}, ignore_index=True)
```
code shown above is for detecting SOG and EOG

Here we send python variable to R interpreter.
**py$** means we want a variable which in python environment.
```{r}
phe_gpp <- read.table("../phenology_new.txt", sep = ',', header = T)
str(phe_gpp)
```

```{r}
phe <- 
phe_new <- merge(Tem, phe_gpp, by.x = c("site","year","month","day"),
                                 by.y = c("site","Year","Month","Day"))
```
```{r}
# fd means frost damages
frostdamage <- phe_new %>% subset(T_min < 0)%>% 
               group_by(site,year)%>%
               summarise(fd = -min(T_min)) 
head(frostdamage)
```
frost damages site-wise
```{r fd,fig.height=20,fig.width=20,fig.cap="Site-wise SPEI distribution",out.height=25,out.width=25, warning=FALSE}
ggplot(frostdamage,aes(factor(year), fd))+geom_point(col="red")+
    facet_wrap(site~.,ncol = 5,scale="free")+
    theme(axis.text.x = element_text(angle = 90))
```
### GPP capacity

GPP capacity is simulated using [**Julia language**](https://julialang.org/) due to it's fast computing.
code as below:

```julia
using Pipe: @pipe # pipeline
using LinearAlgebra
using DataFrames 
using Statistics # basic statistic analysis
using JuliaDB
using Dates # lubridate similar module
using Query # SQL
using LsqFit # linear model
using IterableTables # conver df to tables
using ProgressBars # really bad prograss bar , do not use
using CSV 
using Gadfly 
using SQLite

mydb = SQLite.DB("Flux40SiteDataset.db")
md = DBInterface.execute(mydb, "SELECT * FROM Flux44SiteHalfHourly") |>                 DataFrame

#transform datetime
function Int_2_datetime(x)
          return Dates.DateTime(string(x),"yyyymmddHHMMSS")
       end

Datetime = map(Int_2_datetime, md.TIMESTAMP_END)
md.TIMESTAMP_END = Datetime

# filter by ppfd and Ustar
md = md |> @filter(Dates.month(_.TIMESTAMP_END) >= 4 && 
      Dates.month(_.TIMESTAMP_END) <= 10 && _.PPFD_IN>=4 && _.USTAR>=0.2) |> DataFrame 
      
# drop value = -9999.0     
allowmissing!(md) 
md .= ifelse.(md .== -9999.0, missing, md)
dropmissing!(md)

#################### main function #######################

mydata4 = DataFrame(site = String[],datatime=Date[],Pmax=Float64[],
        Alpha_slope = Float64[],Theta = Float64[],Rd = Float64[],
        R2 = Float64[],GPPcap_2000=Float64[])

sitelist = unique(md.site)

for site in sitelist
    subdata = md |> @filter(_.site == site) |> DataFrame 
    dates = unique([(Date(y)) for y in subdata.TIMESTAMP_END])
    println("now is site--->",site)
    for i in 1:length(dates)  
        d = Date(dates[i]) # go through all dates
        # the windows length is 7 days, and moving step is 1 day
        sdd = subdata |> @filter(_.TIMESTAMP_END >= d &&
             _.TIMESTAMP_END <= d+Dates.Day(7-1))|> DataFrame                    
        gpp = sdd[!,:GPP_NT_VUT_REF] # as a vector with float 64 type  
        light = sdd[!,:PPFD_IN] # vector float 64 type   
        LightCurve(X,p) = @.((p[1]*p[2]*light-
            abs((p[2]*light+p[1])^2-4*p[2]*light*p[3]*p[1])^0.5)/2*p[3] + p[4])
        p0 = [25, 0.001,0.5,2.0] # initial values for parameters as a list     
        fit = curve_fit(LightCurve, light, gpp, p0)
        param = fit.param # get the para list
        yhat = LightCurve(light, fit.param) # fitted values
        R2 = (cor(yhat,gpp))^2  # filter criterion R2 is straitforward 
        # we set Julia day as the middle day of a window
        kk =  (param[2]*2000+param[1])^2-4*param[2]*2000*param[3]*param[1]
        if  kk <= 0
            kk = abs(kk) # if kk < 0 , that means R2 is very very small...
                         # so here it doesn't matter 
        else
            skip
        end
        
        GPPcap_2000 = (param[1]*param[2]*2000-
            (kk)^0.5)/2*param[3] + param[4]
        #append!(GPPcap_2000,LightCurve(X1, fit.param))
        push!(mydata4,(site,d+Dates.Day((7-1)/2),param[1],param[2],param[3],param[4],
                R2,GPPcap_2000))
    end
    println("now the site is finished--->",site)
end

# insert to database
# strang thing is that julia does not supprt inserting dataframe but table type!! so transform to table....

new_data = table(mydata4)
SQLite.load!(new_data,mydb,"GPPcap_2000")

```
for each year in each site, 95% quantile GPP we recognized as yearly GPP capacity.
Strange thing again, the table inserted by Julia can't read by other language, here, for example, R
```{r}
# GPPcap <-
#   dbGetQuery(conn, 'SELECT * FROM GPPcap_2000')
# GPPcap <- data.frame(GPPcap)
GPPcap <- read.table("../GPPcap_2000.txt",sep = ',',header = T)
GPPcap_yearly <- na.omit(GPPcap) %>%
  filter(R2 > 0.6 & GPPcap_2000 > 0) %>%
  group_by(site, year = year(datatime)) %>%
  summarise(Gpp_cap = quantile(GPPcap_2000, 0.95))
```
### Frost vs. GPPcap_2000
```{r}
yearly <- merge(GPPcap_yearly, frostdamage, by = c("site", "year"))
yearly <- merge(
  yearly,
  drought_yearly,
  by.x = c("site", "year"),
  by.y = c("site", "Year")
)

yearly <-
  subset(yearly, !(site %in% c("IT-Ro2", "US-GBT", "US-Me2", "US-UMd")))
#
yearly$fd_nor <- scale(yearly$fd, center = T)
yearly$spei_nor <- scale(yearly$spei, center = T)
yearly$Gpp_cap_nor <- scale(yearly$Gpp_cap, center = T)
# # we exclude 4 sites, dut there are only 4 ~ 5 observations
#
fit1 <- lmerTest::lmer(
  Gpp_cap_nor ~ fd_nor + spei_nor + (1|site ),
  data = yearly,
  REML = F,
  control = lmerControl(optimizer = "optimx",
                        optCtrl = list(method = "nlminb"))
)

parameters(fit1)
```
The above linear mixed model is modified, due to THE variance of random slop for  fd_nor(Normalized frost damage) tends to zero. Thus final model you saw only keeps one random slop and for spei_nor(Normalized SPEI index).
```{r}
summary(fit1)
```

```{r}
randeffect  <- ranef(fit1)$site
randeffect$site <- rownames(randeffect)
head(randeffect)
```
```{r}
# FIX Effect
plot_model(fit1, show.values = TRUE, value.offset = .3,sort.est = TRUE)
```
```{r,fig.width=10}
plot_model(fit1, type = "re",show.values = TRUE, 
           value.offset = .3,sort.est = TRUE)
```
### random effect on geographic distribution
```{r , echo=FALSE}
ggmap::register_google(key = "AIzaSyCqfZYfJ1ZLoktU3BkYHidy4Qye2XRw7LA")
randeffect <- merge(randeffect,forest,by.x = "site",by.y="SITE_ID")
```
Here I hide my private Google map API key... 🤫.
```{r,fig.height=4,fig.width=10,out.height=14,out.width=12,fig.cap="distribution of the forest type of sites"}

qmplot(LOCATION_LONG,LOCATION_LAT,
       data = randeffect,zoom = 4,color = IGBP,size = I(4))+
       scale_color_viridis_d(direction = -1)+
       theme(legend.position = "bottom",legend.title = element_blank())
      
```
The above figure is the distribution of the forest type of sites
```{r,echo=FALSE,fig.height=4,fig.width=8,fig.cap="Random interception distribution"}

# color
pal <- colorNumeric(palette = "PRGn",
                    domain = randeffect$`(Intercept)`)

leaflet(randeffect) %>% addTiles() %>%
  addCircleMarkers(
    data = randeffect,
    lng = ~ LOCATION_LONG,
    lat = ~ LOCATION_LAT,
    stroke = T,
    radius = 6,
    label = ~ site,
    fill = T,
    fillOpacity = 1,
    popup = ~ site,
    color = ~ pal(`(Intercept)`),
     labelOptions = labelOptions(noHide = F,direction = 'auto')
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~ `(Intercept)`,
    title = "Random intercept distribution",
    opacity = 1
  )
```

```{r,echo=FALSE,fig.height=4,fig.width=8,fig.cap="Random slope of SPEI distribution"}
pal <- colorNumeric(palette = "PRGn",
                    domain = randeffect$spei_nor)

leaflet(randeffect) %>% addTiles() %>%
  addCircleMarkers(
    lng = ~ LOCATION_LONG,
    lat = ~ LOCATION_LAT,
    stroke = T,
    radius = 6,
    label = ~ paste(site,":","",round(spei_nor,3)),
    fill = T,
    fillOpacity = 1,
     popup = ~ paste(site,":","",round(spei_nor,3)),
    color = ~ pal(spei_nor),
    labelOptions = labelOptions(noHide = F,direction = 'auto')
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~ spei_nor,
    title = "Random spei slope distribution",
    opacity = 1
  )
```
### different species
```{r}
yearly <-
  merge(yearly,
        select(forest, SITE_ID, IGBP),
        by.x = "site",
        by.y = "SITE_ID")
str(yearly)
```
```{r}
ENF <- subset(yearly,IGBP == 'ENF')

fit2 <- lmerTest::lmer(
  Gpp_cap_nor ~ fd_nor + spei_nor +
    (1 + spei_nor | site),
  data = ENF,
  control = lmerControl(optimizer = "optimx",
                        optCtrl = list(method = "nlminb"))
)

parameters(fit2)
```
```{r}
DBF <- subset(yearly,IGBP == 'DBF')

fit3 <- lmerTest::lmer(
  Gpp_cap_nor ~ fd_nor + spei_nor +
    (1 + spei_nor | site),
  data = DBF,
  control = lmerControl(optimizer = "optimx",
                        optCtrl = list(method = "nlminb")))
parameters(fit3)
```

```{r}
ENF_m <- ENF %>% dplyr::select(-site,-IGBP) %>% group_by(year) %>% summarise_all(mean)
```
```{r}
descdist(ENF_m$Gpp_cap, discrete = FALSE, boot = 1000)
```

```{r}
plot(ENF_m$fd,ENF_m$Gpp_cap)
plot(ENF_m$spei,ENF_m$Gpp_cap)
```
```{r}
basis.spei <-  crossbasis(ENF_m$spei,lag=5, 
                          argvar=list(fun="lin"),
                          arglag=list(fun="poly",degree=2))
basis.GPP <- crossbasis(ENF_m$Gpp_cap, lag=5, argvar=list(fun="lin"),
                      arglag=list(fun="poly",degree=2))
```
```{r}
fitlag <- lm(Gpp_cap ~ basis.spei,data=ENF_m)
```

```{r}
pred1.lag <- crosspred(basis.spei, fitlag, bylag=0.1, cumul=T,at = seq(-1,0,by=0.2))
```
```{r,fig.width=6,fig.height=6}
plot(pred1.lag , "slices", var = -0.6, ci="bars", type="p", col=2, pch=19,
ylab="GPP capacity", ci.level=0.95,cumul = T)

```


